Benchmarks of the Non-Relativistic Hydrodynamics Algorithms in \thornado

### 2017-04-24 Documents/Euler/Benchmarks1D/draft.tex
edits to beginning

\begin{abstract}
The toolkit for high-order neutrino-radiation hydrodynamics (\thornado) is a code for astrophysical radiation-hydrodynamics based on high-order Runge-Kutta Discontinuous Galerkin (RKDG) methods.  
It currently solves the Euler equations for fluid dynamics and the two-moment ($M_{1}$) approximation of the radiative transfer equation. 
In this document, we benchmark the algorithms for solving the Euler equations.  
An extensive set of one-dimensional Riemann problems with known exact solutions were run, and the \thornado\ results were graphed and compared to exact solutions in cases where they exist. 
\end{abstract}

@@

The Euler equations for fluid dynamics \citep[see, e.g.,][for details]{Leveque2002} consists of the mass equation
\begin{equation}
  \frac{\partial \rho}{\partial t}  
  + \frac{\partial}{\partial x} \Big(\rho u\Big) = 0,
  \label{eq:massConservation}
\end{equation}
the momentum equation
\begin{equation}
  \frac{\partial}{\partial t} \Big(\rho u\Big)
  + \frac{\partial}{\partial x} \Big(\rho u^2 + p\Big) = 0,
  \label{eq:momentumConservation}
\end{equation}
and the energy equation
\begin{equation}
  \frac{\partial E}{\partial t}
  + \frac{\partial}{\partial x} \Big((E + p) u\Big) = 0.  
  \label{eq:energyConservation}
\end{equation}
where $\rho$ represents mass density, $u$ the fluid velocity, $p$ the pressure of the fluid, and $E$ the energy.  
Equation~\eqref{eq:massConservation} describes conservation of mass, Equation~\eqref{eq:momentumConservation} describes conservation of momentum, and equation \eqref{eq:energyConservation} describes conservation of energy.
The equation of state
\begin{equation}
  p = (\gamma - 1)
  \Big(E - \frac{1}{2} \rho u^2 \Big)
  \label{eq:eos}
\end{equation}
closes the system of equations, and links the pressure to the quantities evolved in Equations \eqref{eq:massConservation}-\eqref{eq:energyConservation}. 
The ratio of specific heats, $\gamma$, is taken to be a constant.

@@

The computational domain $D=\{x:x\in[x_{\mbox{\tiny L}},x_{\mbox{\tiny R}}]\}$ is subdivided into $N$ elements $I_{i}=\{x:x\in(x_{i-1/2},x_{i+1/2})\}$.  
The width of element $I_{i}$ is then $\Delta x_{i}=x_{i+1/2}-x_{i-1/2}$.  
The semi-discrete DG method is then to find $\boldsymbol{U}_{h}$ such that
\begin{equation}
  \partial_{t}\int_{I_{i}}\boldsymbol{U}_{h}\,\phi_{j}\,dx
  +\big(\widehat{\boldsymbol{F}}\,\phi_{j}|_{i+1/2}-\widehat{\boldsymbol{F}}\,\phi_{j}|_{i-1/2}\big)
  -\int_{I_{i}}\boldsymbol{F}(\boldsymbol{U}_{h})\,\partial_{x}\phi_{j}\,dx=0
  \label{eq:semidiscreteDG}
\end{equation}
for all $\phi_{j}$ and all $I_{i}\in D$.  
In Equation~\eqref{eq:semidiscreteDG}, $\widehat{\boldsymbol{F}}_{i\pm1/2}$ is a numerical flux approximating the flux across the interfaces connecting neighboring elements.  
In \thornado, we have implemented the Local Lax-Friedrichs (LLF), the Harten-Lax-van Leer (HLL), and the HLLC Riemann solvers, but in the numerical experiments presented here, we only use the HLLC solver.  
To complete the spatial discretization, we use Gaussian quadratures to evaluate the integrals in Eq.~\eqref{eq:semidiscreteDG}.  
During time-integration, we apply limiters to the polynomial representation.  
We use the TVB limiter of \citet{} to prevent oscillations near discontinuities and the positivity limiter of \citet{ZhangShu2010} to ensure that the density and pressure remain positive.

We apply the method of lines approach to temporal discretization.  
With the spatial discretization specified, the DG problem in Eq.~\eqref{eq:semidiscreteDG} can be viewed as a system of ordinary differential equations (ODEs) for the expansion coefficients $\boldsymbol{C}=\{\boldsymbol{C}_{j}\}_{j=0}^{k}$, i.e.,
\begin{equation}
  d_{t}\boldsymbol{C}
  =\boldsymbol{L}(\boldsymbol{C}),
\end{equation}
which can be discretized in time.  
To this end, we apply the strong stability-preserving Runge-Kutta (SSP-RK) methods of order 1 to 3 \citep{gottlieb_etal_2001}.  
The 1st order method (SSP-RK1) is simply the forward Euler method, while the 2nd and 3rd order methods (SSP-RK2 and SSP-RK3, respectively) can be expressed as convex combinations of forward Euler steps.  
A general $m$-stage Runge-Kutta method is given by
\begin{align}
  \boldsymbol{C}^{(0)}
  &=\boldsymbol{C}^{n} \\
  \boldsymbol{C}^{(i)}
  &=\sum_{l=1}^{i-1}\alpha_{il}\,\big(\,\boldsymbol{C}^{(l)}+\beta_{il}\,\Delta t\,\boldsymbol{L}(\boldsymbol{C}^{(l)})\,\big)
  \quad i=1,\ldots,m \\
  \boldsymbol{C}^{n+1}
  &=\boldsymbol{C}^{(m)},
\end{align}
where $\alpha_{il},\beta_{il}\ge0$ are given by Butcher tables.  
During time integration, we apply slope and positivity limiters after each stage as a post-processing step.

@@

The code was tested using Riemann problems. The Riemann problem models a fluid in an area of one-dimensional space. The initial data features one jump discontinuity, and is constant on either side of this discontinuity. The computational domain is divided into space and time cells. The conditions of each cell are advanced forward for all space cells, one time step at a time, according to the Euler equations for fluid dynamics. \citep{Leveque2002}

@@

Several tests were used to assess the accuracy of the integration method \citep{LiskaWendroff2003}. 
\begin{table}[h!]
\begin{center}
\begin{tabular}{ |c|c c c|c c c|c c| }
 \hline
 Test & $\rho_L$ & $u_L$ & $p_L$ & $\rho_R$ & $u_R$ & $p_R$ & $x_0$ & $T$ \\
 \hline
 1 & 1 & 0 & 1 & 0.125 & 0 & 0.1 & 0.5 & 0.2 \\
 2 & 1 & 0.75 & 1 & 0.125 & 0 & 0.1 & 0.3 & 0.2 \\
 3 & 1 & -2 & 0.4 & 1 & 2 & 0.4 & 0.5 & 0.15 \\
 4 & 1 & 1 & $10^{-6}$ & 1 & -1 & $10^{-6}$ & 0.5 & 1 \\
 5 & 1 & -19.59745 & 1000 & 1 & -19.59745 & 0.01 & 0.8 & 0.012 \\
 6 & 5.99924 & 19.5975 & 460.894 & 5.99242 & -6.19633 & 46.095 & 0.4 & 0.035 \\
 7 & 1.4 & 0 & 1 & 1 & 0 & 1 & 0.5 & 2 \\
 8 & 1.4 & 0.1 & 1 & 1 & 0.1 & 1 & 0.5 & 2 \\
 9 & 0.1261192 & 8.9047029 & 782.92899 & 6.591493 & 2.2654207 & 3.1544874 & 0.5 & 0.0039 \\
10 & 5.6698 & 9.0299 & 100 & 1.0 & 0.0 & 1.0 & 0 & 1 \\
11 & 5.6698 & -1.4701 & 100 & 1.0 & -10.5 & 1.0 & 0 & 1 \\
 \hline
\end{tabular}
  \caption{Initial parameters for most of the tests described in this paper, from \citep{LiskaWendroff2003}~and \citep{Leveque2002}. The table shows initial density ($\rho$), velocity ($u$), and pressure ($p$) on both sides of a discontinuity located at $x_0$. These initial conditions are then advanced to time $T$.}
  \label{tab:testtable}
\end{center}
\end{table}
$x_0$ refers to the position of the barrier. 
The computational domain $D$ is defined with $x_{\mbox{\tiny L}}=0.0$ and $x_{\mbox{\tiny R}}=1.0$ in tests 1--8, $x_{\mbox{\tiny L}}=0.1$ and $x_{\mbox{\tiny R}}=0.6$ in test 9 \citep{LiskaWendroff2003}, $x_{\mbox{\tiny L}}=-5$ and $x_{\mbox{\tiny R}}=15$ in test 10, and $x_{\mbox{\tiny L}}=-15$ and $x_{\mbox{\tiny R}}=5$ in test 11 \citep{Leveque2002}. 
\ee{Provide more detailed description of the Riemann problems and refer to Table~\ref{tab:testtable}}

>>>>

\begin{abstract}
The toolkit for high-order neutrino-radiation hydrodynamics (\thornado) is a code for astrophysical radiation-hydrodynamics based on high-order Runge-Kutta Discontinuous Galerkin (RKDG) methods.  
It currently solves the Euler equations for fluid dynamics and the two-moment ($M_{1}$) approximation of the radiative transfer equation, both in the non-relativistic limits. 
In this document, we benchmark the algorithms for solving the non-relativistic Euler equations.  
Results from an extensive set of one-dimensional Riemann problems are compared to exact solutions in cases where they exist. We also show results for th interacting blast wave problem of Colella and Woodward and the shock entropy wave problem of Shu and Oster. 
\end{abstract}

@@

The Euler equations for fluid dynamics \citep[see, e.g.,][for details]{Leveque2002} consists of the mass conservation equation
\begin{equation}
  \frac{\partial \rho}{\partial t}  
  + \frac{\partial}{\partial x} \Big(\rho u\Big) = 0,
  \label{eq:massConservation}
\end{equation}
the momentum conservation equation
\begin{equation}
  \frac{\partial}{\partial t} \Big(\rho u\Big)
  + \frac{\partial}{\partial x} \Big(\rho u^2 + p\Big) = 0,
  \label{eq:momentumConservation}
\end{equation}
and the energy conservation equation
\begin{equation}
  \frac{\partial E}{\partial t}
  + \frac{\partial}{\partial x} \Big((E + p) u\Big) = 0.  
  \label{eq:energyConservation}
\end{equation}
where $\rho$ represents mass density, $u$ the fluid velocity, $p$ the fluid pressure, and $E=e+\frac{1}{2}\rho u^2$ the total energy (internal plus kinetic).  
Equation~\eqref{eq:massConservation} expresses conservation of mass, Equation~\eqref{eq:momentumConservation} expresses conservation of momentum, and equation \eqref{eq:energyConservation} expresses conservation of energy.
The equation of state
\begin{equation}
  p = (\gamma - 1)
  \Big(E - \frac{1}{2} \rho u^2 \Big)
  \label{eq:eos}
\end{equation}
closes the system of equations, and links the pressure to the quantities evolved by Equations \eqref{eq:massConservation}-\eqref{eq:energyConservation}. 
The ratio of specific heats, $\gamma$, is taken to be a constant.

@@

The computational domain $D=\{x:x\in[x_{\mbox{\tiny L}},x_{\mbox{\tiny R}}]\}$ is subdivided into $N$ elements $I_{i}=\{x:x\in(x_{i-1/2},x_{i+1/2})\}$ for $i=1,\ldots,N$.  
The width of element $I_{i}$ is then $\Delta x_{i}=x_{i+1/2}-x_{i-1/2}$.  
The semi-discrete DG method is then to find $\boldsymbol{U}_{h}$ such that
\begin{equation}
  \partial_{t}\int_{I_{i}}\boldsymbol{U}_{h}\,\phi_{j}\,dx
  +\big(\widehat{\boldsymbol{F}}\,\phi_{j}|_{i+1/2}-\widehat{\boldsymbol{F}}\,\phi_{j}|_{i-1/2}\big)
  -\int_{I_{i}}\boldsymbol{F}(\boldsymbol{U}_{h})\,\partial_{x}\phi_{j}\,dx=0
  \label{eq:semidiscreteDG}
\end{equation}
for all $\phi_{j}$ and all $I_{i}\in D$.  
In Equation~\eqref{eq:semidiscreteDG}, $\widehat{\boldsymbol{F}}_{i\pm1/2}$ is a numerical flux approximating the flux across the interfaces connecting neighboring elements.  
In \thornado, we have implemented the Local Lax-Friedrichs (LLF), the Harten-Lax-van Leer (HLL), and the HLLC Riemann solvers, but in the numerical experiments presented here, we only use the HLLC solver.  
To complete the spatial discretization, we use Gaussian quadratures to evaluate the integrals in Eq.~\eqref{eq:semidiscreteDG}.

We apply the method of lines approach to temporal discretization.  
With the spatial discretization specified, the DG problem in Eq.~\eqref{eq:semidiscreteDG} can be viewed as a system of ordinary differential equations (ODEs) for the expansion coefficients $\boldsymbol{C}=\{\boldsymbol{C}_{j}\}_{j=0}^{k}$, i.e.,
\begin{equation}
  d_{t}\boldsymbol{C}
  =\boldsymbol{L}(\boldsymbol{C}),
\end{equation}
which can be discretized in time.  
To this end, we apply the strong stability-preserving Runge-Kutta (SSP-RK) methods of order 1 to 3 \citep{gottlieb_etal_2001}.  
The 1st order method (SSP-RK1) is simply the forward Euler method, while the 2nd and 3rd order methods (SSP-RK2 and SSP-RK3, respectively) can be expressed as convex combinations of forward Euler steps.  
A general $m$-stage Runge-Kutta method is given by
\begin{align}
  \boldsymbol{C}^{(0)}
  &=\boldsymbol{C}^{n} \\
  \boldsymbol{C}^{(i)}
  &=\sum_{l=1}^{i-1}\alpha_{il}\,\big(\,\boldsymbol{C}^{(l)}+\beta_{il}\,\Delta t\,\boldsymbol{L}(\boldsymbol{C}^{(l)})\,\big)
  \quad i=1,\ldots,m \\
  \boldsymbol{C}^{n+1}
  &=\boldsymbol{C}^{(m)},
\end{align}
where $\alpha_{il},\beta_{il}\ge0$ are given by Butcher tables. During time-integration, we apply limiters to the polynomial representation after each stage as a post-processing step.  
We use the TVB limiter of \citet{} to prevent oscillations near discontinuities and the positivity limiter of \citet{ZhangShu2010} to ensure that the density and pressure remain positive.

@@

The code was tested against a series of one-dimensional Riemann problems. The Riemann problem models a fluid in an area of one-dimensional space. The initial data features one jump discontinuity, and is constant on either side of this discontinuity. The computational domain is divided into space and time cells. The conditions of each cell are advanced forward for all space cells, one time step at a time, according to the Euler equations for fluid dynamics. \citep{Leveque2002}

@@

Several tests were used to assess the accuracy of the RKDG method implemented in thornado \citep{LiskaWendroff2003}. 
\begin{table}[h!]
\begin{center}
\begin{tabular}{ |c|c c c|c c c|c c| }
 \hline
 Test & $\rho_L$ & $u_L$ & $p_L$ & $\rho_R$ & $u_R$ & $p_R$ & $x_0$ & $T$ \\
 \hline
 1 & 1 & 0 & 1 & 0.125 & 0 & 0.1 & 0.5 & 0.2 \\
 2 & 1 & 0.75 & 1 & 0.125 & 0 & 0.1 & 0.3 & 0.2 \\
 3 & 1 & -2 & 0.4 & 1 & 2 & 0.4 & 0.5 & 0.15 \\
 4 & 1 & 1 & $10^{-6}$ & 1 & -1 & $10^{-6}$ & 0.5 & 1 \\
 5 & 1 & -19.59745 & 1000 & 1 & -19.59745 & 0.01 & 0.8 & 0.012 \\
 6 & 5.99924 & 19.5975 & 460.894 & 5.99242 & -6.19633 & 46.095 & 0.4 & 0.035 \\
 7 & 1.4 & 0 & 1 & 1 & 0 & 1 & 0.5 & 2 \\
 8 & 1.4 & 0.1 & 1 & 1 & 0.1 & 1 & 0.5 & 2 \\
 9 & 0.1261192 & 8.9047029 & 782.92899 & 6.591493 & 2.2654207 & 3.1544874 & 0.5 & 0.0039 \\
10 & 5.6698 & 9.0299 & 100 & 1.0 & 0.0 & 1.0 & 0 & 1 \\
11 & 5.6698 & -1.4701 & 100 & 1.0 & -10.5 & 1.0 & 0 & 1 \\
 \hline
\end{tabular}
  \caption{Initial parameters for most of the tests described in this paper, from \citep{LiskaWendroff2003}~and \citep{Leveque2002}. The table shows initial density ($\rho$), velocity ($u$), and pressure ($p$) on both sides of a discontinuity located at $x_0$. These initial conditions are then advanced to time $T$.}
  \label{tab:testtable}
\end{center}
\end{table}
TAble 1 dispplays the initial data for the Riemann problems. ($\rho_L$, $u_L$ , $p_L$) denote the mass density, velocity, and pressure to the left of the interface at $x=x_0$;($\rho_R$, $u_R$,  $p_R$) refer to these same quantities to the right of the interface. The problem is advanced to time T.
The computational domain $D$ is defined with $x_{\mbox{\tiny L}}=0.0$ and $x_{\mbox{\tiny R}}=1.0$ in tests 1--8, $x_{\mbox{\tiny L}}=0.1$ and $x_{\mbox{\tiny R}}=0.6$ in test 9 \citep{LiskaWendroff2003}, $x_{\mbox{\tiny L}}=-5$ and $x_{\mbox{\tiny R}}=15$ in test 10, and $x_{\mbox{\tiny L}}=-15$ and $x_{\mbox{\tiny R}}=5$ in test 11 \citep{Leveque2002}. 
%\ee{Provide more detailed description of the Riemann problems and refer to Table~\ref{tab:testtable}}
